Grade Retention and Mixed Race

### 2022-08-18 products/paper/main.qmd
Add changes from first editing of into and part of background

The experiences of mixed race individuals will play an important role in the future of race in the US. A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines (==CITE==). This demographic phenomenon may have a potentially transformative effect on racial boundaries. Although race mixing has always been a feature of US history (and the history of any racialized social system) (==CITE==), this biracial boom is notable for its demographic size and for occurring in  post-Civil Rights ear in which historical norms govern how to handle racial mixture may be dissolving or shifting. Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into the prevailing racial hierarchy that reinforces a system of racial inequality.

Extensive scholarship documents the identification and classification decisions of multiracial individuals (==CITE==). ==SUMMARIZE BRIEFLY==. On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers. The lack of work on this topic, despite its acknowledged relevance, results from two important limitations in most available data sources.

First, using direct identification of biracial respondents on data sources such as the Census or large scale surveys is problematic. Research on the identification and classification of biracial individuals indicates that many individuals whose parents belong to different races will not always or consistently identify with both races (==CITE==). Social status and social mobility has been shown to predict racial identification (==CITE==) and this process of reverse causation is likely to be strongest for more ambiguous groups such as biracials. Thus, differences in outcome may in some part produce differences in identification rather than the other way around when examining biracial individuals identified by self-classification.

@@

In this article, I exploit features of the American Community Survey to identify one life outcome that is available in large scale Census data - grade retention. Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race (==CITATIONS==). Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience in their life and within it we can clearly see the imprint of the racial hierarchy. Because, this outcome occurs to children living in an often natal household, I can identify the race of both biological parents for many cases. This identification allows me to compare the likelihood of grade retention for biracial children in comparison to their monoracial peers. Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents.

@@

The most direct pathway for understanding the placement of this outcome is via how biracial individuals are classified by those in positions of power and control, such as teachers, administrators, and employers. To the extent that biracial individuals are classified as more like one of their monoracial constituent groups. they will receive the same treatment and either reap similar rewards or suffer similar discrimination and penalties.

Two well known norms have long governed such patterns of identification. The pattern of *hypodescent* dictates that individuals with mixed racial ancestry will be classified as members of the lower status group while the pattern of *hyperdescent* dictates that such individuals will be classified as members of the higher status  (==CITATIONS==) Hypodescent, in the form of the "one drop rule" has long characterized the experiences of mixed part-Black individuals who are treated as exclusively Black (==CITE F JAMES DAVIS==). The pattern of hyperdescent blends mixed race individuals into the often unmarked higher status group and is thus more difficult to see in practice. Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming to be White (==CITATIONS==). ==NEED TO MIX IN SOME OF THE COMMENTS IN ANN AND ALIYAS RECENT PAPER==

@@

Notably, both of these mechanisms could co-exist with norms of hypo/hyperdescent. We know, for example, that lighter skin tone among monoracial Black individuals is associated with better outcomes (==CITATION==). So, while the one-drop rule provides a consistent classification system, it does not create uniform experiences for all Blacks and skin tone may help to "soften" the experience of discrimination. Similarly, if an in-between status results for biracial individuals becaise they are less consistently classified, that does not mean that the classification system and the norms that underlie it are weakening. (==CITE MONK HERE==)

@@

All of these arguments relate to where we might expect the outcomes of biracial individuals to fall relative to their monoracial peers. However, the experience of hybridity itself may have positive and negative effects that do not tell us the location itself but rather may modify it in the context of other patterns. On the positive side, hybridity may enlarge the cultural "toolkit" (==CITATION==) of biracial individuals. This enlarged toolkit may allow such individuals to more easily navigate a variety of social situations (e.g. code switch). On the negative side, Stonequist's (==CITATION==) "marginal man" hypothesis suggests that the dissonance of an identity that crosses strong and salient racial boundaries will be experienced by biracial individuals as not fitting in anywhere. This liminality may result in negative mental health experiences for biracial individuals that may affect other types of outcomes as well.

@@

In addition to these specifically racial mechanisms that may drive the placement of biracial individuals, researchers must also pay careful attention to how socioeconomic background may affect the placement of biracial individuals within a racial hierarchy. Socioeconomic background plays a role in racial inequality more generally as we know that accounting for socioeconomic difference often reduces the magnitude of observed racial inequality, suggesting that some of racial inequality we observe operates through class differences (CITATION). This issue is particularly important for biracial individuals because of potential class selection into interracial unions. The likelihood of being an interracial union varies by class background for at least some groups (CITATION), although it varies in complex ways and we know more about interracial marriages than other types of unions that may produce children. For example, black partners tend to be positively selected with regard to education into Black/White marriages while white partners are negatively selected (CITATION). According to status exchange theory (CITATIONS), this pattern may be common to all types of interracial marriages within an unequal status hierarchy, although the evidence of status exchange for other forms of marriages is weaker (CITATION). How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena for a given pairing, the gender balance in that pairing, and how the gender of parents affects outcomes.

>>>>

The experiences of mixed race individuals will play an important role in the future of race in the US. A rise in interracial marriage and coupling since the 1960s has resulted in a "biracial baby boom" of individuals whose parents cross racial lines (==CITE==). This demographic phenomenon may have a potentially transformative effect on racial boundaries. Although race mixing has always been a feature of US history (==CITE==), this biracial boom is notable for its demographic size and for occurring in a post-Civil Rights era in which historical norms governing how to handle racial mixture may be shifting. Understanding the transformative potential of this biracial baby boom is contingent on developing a better understanding of how these biracial individuals identify their own race and how they are sorted into a prevailing racial hierarchy that reinforces a system of racial inequality.

Extensive scholarship documents the identification and classification decisions of multiracial individuals (==CITE==). ==SUMMARIZE BRIEFLY==. On the other hand, only a smattering of research has documented how the life chances of biracial individuals compare to their monoracial peers. (==CITATION==) The lack of work on this topic, despite its acknowledged relevance, results from two important limitations in most available data sources.

First, using direct identification of biracial respondents on data sources such as the Census or large scale surveys is problematic. Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races will not always or consistently identify with both races (==CITE==). Furthermore, social status and social mobility has been shown to predict racial identification (==CITE==) and this process of reverse causation is likely to be strongest for more ambiguous groups such as biracials. Thus, differences in outcome may in some part produce differences in identification rather than the other way around when examining biracial individuals identified by self-identification.

@@

When children reside with both biological parents, biracial individuals can be identified in large scale data sources such as the Census. Researchers have long utilized the ability to identify such biracial children to examine the classification choices made on behalf of these children (==CITATIONS==). Because of their age, it is more difficult to analyze the outcomes of such children due to the simple fact that they haven't had many measurable life outcomes yet.

However, one important outcome, grade retention, is measurable and important even for young children. Being held back in school is a relatively common occurrence that has consequences for later educational transitions and is disproportionately applied by race (==CITATIONS==). Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience and within it we can clearly see the imprint of the racial hierarchy. In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers. Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents.

@@

The most direct pathway for such placement is via how biracial individuals are classified by those in positions of power and control, such as teachers, administrators, and employers. To the extent that biracial individuals are classified as more like one of their monoracial constituent groups. they will receive the same treatment and either reap similar rewards or suffer similar discrimination and penalties.

Two well known norms have long governed such patterns of classification. The pattern of *hypodescent* dictates that individuals with mixed racial ancestry will be classified as members of the lower status group while the pattern of *hyperdescent* dictates that such individuals will be classified as members of the higher status group (==CITATIONS==). Hypodescent, in the form of the "one drop rule" has long characterized the experiences of mixed part-Black individuals who are treated as exclusively Black (==CITE F JAMES DAVIS==). The pattern of hyperdescent blends mixed race individuals into the often unmarked higher status group and is thus more difficult to see in practice. Indigenous ancestry (American Indian, Pacific Islander, and Alaska Native) has historically followed a logic of hyperdescent in which a mixing of indigenous ancestry does not disqualify individuals from claiming to be White (==CITATIONS==). ==NEED TO MIX IN SOME OF THE COMMENTS IN ANN AND ALIYAS RECENT PAPER==

@@

Notably, both of these mechanisms could co-exist with norms of hypo/hyperdescent. We know, for example, that lighter skin tone among monoracial Black individuals is associated with better outcomes (==CITATION==). So, while the one-drop rule provides a consistent classification system, it does not create uniform experiences for all Blacks and skin tone may help to "soften" the experience of discrimination. (==CITE MONK HERE==)

@@

The experience of hybridity itself may also positively and/or negative affect the outcomes for biracial individuals. On the positive side, hybridity may enlarge the cultural "toolkit" (==CITATION==) of biracial individuals. This enlarged toolkit may allow such individuals to more easily navigate a variety of social situations (e.g. code switch). On the negative side, Stonequist's (==CITATION==) "marginal man" hypothesis suggests that the dissonance of an identity that crosses strong and salient racial boundaries will be experienced by biracial individuals as not fitting in anywhere. This liminality may result in negative mental health experiences for biracial individuals that can then affect other types of outcomes as well.

@@

In addition to the aforementioned racial mechanisms that may drive the placement of biracial individuals, researchers must also pay careful attention to how socioeconomic background may affect the placement of biracial individuals within a racial hierarchy. Socioeconomic background plays a role in racial inequality more generally as we know that accounting for socioeconomic difference often reduces the magnitude of observed racial inequality, suggesting that some of the racial inequality we observe operates through class differences (CITATION). This issue is particularly important for biracial individuals because of potential class selection into interracial unions. The likelihood of being an interracial union varies by class background for at least some groups (CITATION), although it varies in complex ways and we know more about interracial marriages than other types of unions that may produce children. For example, black partners tend to be positively selected with regard to education into Black/White marriages while white partners are negatively selected (CITATION). According to status exchange theory (CITATIONS), this pattern may be common to all types of interracial marriages within an unequal status hierarchy, although the evidence of status exchange for other forms of marriages is weaker (CITATION). How this affects the relative resources of biracial individuals will depend then in some part on the relative strength of these phenomena for a given pairing, the gender balance in that pairing, and how the gender of parents affects outcomes.
